\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2016-06-28 paper/ms.tex
Finished comparison section

\section{Relation to Other Algorithms}\label{sec:relation}
The simple C implementation of Nested Sampling given by
\citet{skilling2006nested}

\pkg{MultiNest} \citep{feroz2009multinest} is a Nested Sampling implementation
that dispenses with MCMC for generating new particles from constrained
prior distributions. Instead, the new particles are generated by making
an approximation to the constrained prior using ellipses. This tends to work
well in low to moderate dimensional parameter spaces (up to $\sim 50$).
In higher dimensions, \pkg{DNest4} is more useful than \pkg{MultiNest}
since it is based on MCMC. We expect \pkg{MultiNest} to be more useful
than
\pkg{DNest4} only on low to moderate dimensional problems with slow
likelihood evaluations.

\pkg{emcee} \citep{emcee} is a popular \proglang{Python} package for MCMC
based on affine-invariant ensemble sampling.

The popular \pkg{JAGS} \citep{jags} and \pkg{Stan} \citep{stan} packages
are more convenient than \pkg{DNest4} for specifying models, by virtue
of their model-specification languages. However, being based on Gibbs
sampling and Hamiltonian MCMC respectively, they can run into difficulty
on multimodal posterior distributions. Also, they do not calculate
marginal likelihoods.

The advantages and disadvantages of these various packages, in the opinion
of the authors, are summarized in Table~\ref{tab:packages}.

@@

\begin{table}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Package     &   Easy to     & High   &   Multimodal     & Dependent   &   Phase   &   Computes $Z$?\\
            &   implement & dimensions? & distributions? & distributions? & changes? & \\
& models? & & & & &\\
\hline
\pkg{DNest4}    & \xmark  & \cmark & \cmark & \cmark &\cmark &\cmark\\
\pkg{emcee}    & \cmark  & \xmark & \xmark & \cmark &\xmark &\xmark\\
\pkg{JAGS}    & \cmark  & \cmark & \xmark & \xmark & \xmark &\xmark\\
\pkg{MultiNest}    & \cmark  & \xmark & \cmark & \cmark & \cmark &\cmark\\
\pkg{Stan}    & \cmark  & \cmark & \xmark & \cmark & \xmark &\xmark\\
\hline
\end{tabular}}
\caption{\label{tab:packages}}
\end{center}
\end{table}

>>>>

\section{Relation to Other Algorithms}\label{sec:relation}
The Diffusive Nested Sampling algorithm, and its implementation in
\pkg{DNest4}, has advantages and disadvantages compared to other Bayesian
computation algorithms and software.

In the Nested Sampling sphere,
the simple \proglang{C} implementation of given by
\citet{skilling2006nested} is useful for understanding the classic NS
algorithm. Similarly to \pkg{DNest4}, it is the user's responsibility
to implement the MCMC moves used. The DNS algorithm produces more
accurate estimates of $Z$ and ought to outperform classic NS on multimodal
problems.

\pkg{MultiNest} \citep{feroz2009multinest} is a
\proglang{FORTRAN}
Nested Sampling implementation
that dispenses with MCMC for generating new particles from constrained
prior distributions. Instead, the new particles are generated by making
an approximation to the constrained prior using ellipsoids. This tends to work
well in low to moderate dimensional parameter spaces (up to $\sim 40$).
In higher dimensions, \pkg{DNest4} is more useful than \pkg{MultiNest}
since it is based on MCMC. We expect \pkg{MultiNest} to be more useful
than
\pkg{DNest4} only on low to moderate dimensional problems with slow
likelihood evaluations, where \pkg{DNest4}'s reliance on MCMC becomes costly.
The more recent \pkg{POLYCHORD} \citep{handley2015polychord} combines
\pkg{MultiNest}-like methods with MCMC.

Outside of Nested Sampling,
the popular \pkg{JAGS} \citep{jags} and \pkg{Stan} \citep{stan} packages
are more convenient than \pkg{DNest4} for specifying models, by virtue
of their model-specification languages. For users who are interested only
in the posterior distribution and do not need the marginal likelihood,
these are very useful.
However, being based on Gibbs
sampling and Hamiltonian MCMC respectively, they can run into difficulty
on multimodal posterior distributions.

\pkg{emcee} \citep{emcee} is a popular \proglang{Python} package for MCMC
based on affine-invariant ensemble sampling. The user needs only to
specify a \proglang{Python} function evaluating the log of the posterior
density (up to a normalizing constant). This user-friendliness is a key
advantage of \pkg{emcee}, and its algorithm performs well on low--moderate
dimensional (up to $\sim 50$) parameter spaces with a unimodel
target distribution (which can be highly dependent). However, it
can give misleading results in higher dimensions \citep{huijser}.

Algorithms based on ``annealing'' or ``tempering'', such as parallel
tempering \citep{hansmann1997parallel} and annealed importance sampling
\citep{neal2001annealed} are
related to Nested Sampling and are useful on multimodal posterior distributions.
However, they require much more tuning than NS (in the form of an
{\em annealing schedule}) and do
not work on phase change problems, unlike Nested Sampling
\citep{skilling2006nested}.

The advantages and disadvantages of a subset of these
packages are summarized in Table~\ref{tab:packages}.

@@

\begin{table}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Package     &   Easy to     & High   &   Multimodal     & Dependent   &   Phase   &   Computes $Z$?\\
            &   implement & dimensions? & distributions? & distributions? & changes? & \\
& models? & & & & &\\
\hline
\pkg{DNest4}    & \xmark  & \cmark & \cmark & \cmark &\cmark &\cmark\\
\pkg{emcee}    & \cmark  & \xmark & \xmark & \cmark &\xmark &\xmark\\
\pkg{JAGS}    & \cmark  & \cmark & \xmark & \xmark & \xmark &\xmark\\
\pkg{MultiNest}    & \cmark  & \xmark & \cmark & \cmark & \cmark &\cmark\\
\pkg{Stan}    & \cmark  & \cmark & \xmark & \cmark & \xmark &\xmark\\
\hline
\end{tabular}}
\caption{A simplified summary of the advantages and disadvantages of some
Bayesian computation software packages.\label{tab:packages}}
\end{center}
\end{table}
